The Precambrian paleogeography of Laurentia

### 2020-11-11 Manuscript/Laurentia_Paleogeo_Manuscript.tex
update figure numbers and add NW Laurentia rifting text

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Tectonic_history.pdf}
\caption{\textbf{Simplified timeline of Laurentia's tectonic history over the past $\sim$1.8 billion years.} Brief summaries and references related to the orogenic and rifting episodes are given in the text. A timeline of large igneous provinces (LIPs) associated with typically brief and voluminous (or interpreted to be voluminous) volcanism is also shown. The interpreted age of paleomagnetic poles for Laurentia (not including separated terranes) compiled in this study for the Proterozoic and in \cite{Torsvik2012a} for the Phanerozoic is shown. Abbreviations on the figure: CAMP -- Central Atlantic Magmatic Province; proto-Cord -- proto-Cordilleran.}
\label{fig:tectonic_history}
\end{figure*}

@@

In northwest Laurentia from the Ogilvie Mountains of Yukon to Victoria Island, the sedimentary rock record is distinct from further south as it also records earlier Neoproterozoic basin development during the Tonian Period in addition to Cryogenian basin development \citep{Macdonald2012a}. Lithospheric extension is interpreted from basin development that accommodated deposition of the lower Fifteenmile Group with maximum depositional ages of ca. 1050 Ma with ongoing basin development ca. 812 Ma (age constraint from a U-Pb zircon date on a tuff within the upper Fifteenmile Group; \citealp{Macdonald2010a}) potentially through thermal subsidence \citep{Macdonald2012a}. Earlier basin development in the region recorded by the Mesoproterozoic/Neoproterozoic Pinguicula Group could provide valuable insight on tectonic history as it has been interpreted to have been deposited in an extensional basin \citep{Medig2016a}, however it is poorly constrained in terms of age --- older than the Fifteenmile Group and younger than the ca. 1382 Ma Hart River sills (which themselves have been interpreted to be emplaced in conjunction with rifting; \citealp{Verbaas2018a}).

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Laurentia_poles_combined.pdf}
\caption{\textbf{Paleomagnetic poles from Laurentia.} Upper-left panel: Paleomagnetic poles from 1800 to 720 Ma for Laurentia (including Greenland). Portions of the apparent polar wander path (APWP) are referred to by the names Logan loop, Keweenawan track, and Grenville loop in the literature and those are labeled next to the associated poles. Upper-right panel:  Paleomagnetic poles from 1800 to 580 Ma for Laurentia (including those from the separated terranes of Greenland, Scotland and Svalbard rotated to Laurentia coordinates). The youngest poles from the Ediacaran Period have unusually variable positions as discussed in the text. Lower-left panel: Running mean APWP calculated with a 20 million year moving window. Lower-right panel: Poles for the Archean provinces of Laurentia prior to Laurentia's Paleoproterozoic amalgamation.}
\label{fig:Laurentia_poles}
\end{figure*}

@@

%The poles on this figure should be labeled
\begin{figure*}
\centering
\includegraphics[width=6 in]{../Figures/Superior_Slave_reconstructions.pdf}
\caption{\textbf{Paleogeographic reconstructions developed using poles from the Superior, Slave and Rae provinces.} The polarity options that are chosen for the provinces are those that minimize total apparent polar wander path length. This model reconstructs a wide Manikewan ocean that underwent orthogonal closure rather than an alternative possibility of a narrower Manikewan ocean with a pivot-like closure. Paleomagnetic poles are shown colored to match their respective province with these provinces shown in present-day coordinates and labeled in the 0 Ma panel. Poles with ages that are within 25 million years of the given time slice are shown. The relatively well-resolved pole paths from the Superior and Slave provinces (Fig. \ref{fig:Laurentia_poles}) that are utilized for these reconstructions provide strong support for differential plate tectonic motion between 2220 and 1850 Ma.}
\label{fig:Superior_Slave_recons}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=0.9\textwidth]{../Figures/Laurentia_paleolatitude.pdf}
\caption{\textbf{Laurentia paleolatitude through time in data and models.} Top panel: Paleolatitude for the city of Duluth on the southern margin of the Superior province (present-day coordinates of lat=46.79\textdegree N, lon=92.10\textdegree W) implied by paleomagnetic poles from Laurentia and associated terranes. The paleomagnetic poles are compiled in Table 2. Middle panel: Paleolatitude implied by Laurentia poles compared with that implied by published paleogeographic models and the simple Laurentia model used in this chapter for the reconstructions in Figure \ref{fig:Laurentia_reconstructions}. TC2017 refers to \cite{Torsvik2017a} and S-H 2017 refers to \cite{Swanson-Hysell2017a}. Bottom panel: the velocity implied by the continuous paleogeographic models in cm per year for the Duluth reference point on Laurentia.}
\label{fig:Laurentia_paleolatitude}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Laurentia_reconstructions.pdf}
\caption{\textbf{Paleogeographic reconstructions of Laurentia at time intervals through the Proterozoic.} These reconstructions use the simple Laurentia pole interpolation model that is shown in Figure \ref{fig:Laurentia_paleolatitude} to reconstruct the tectonic elements of \cite{Whitmeyer2007a} shown in Figure \ref{fig:Laurentia_map}. Modern coastlines are maintained in these polygons so that the rotated orientations can be interpreted by the reader in comparison to Figure \ref{fig:Laurentia_map}. Paleomagnetic poles within 25 million years of each reconstruction time are plotted. All reconstructions have poles within such a time frame that provide constraints with the exception of the 850 Ma reconstruction which is shown faintly given this relative uncertainty in Laurentia's position.}
\label{fig:Laurentia_reconstructions}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=7 in]{../Figures/Laurentia_evaporite_figure.pdf}
\caption{\textbf{Paleolatitude of Laurentia evaporites.} Left panel: The paleolatitude of evaporite deposits as reconstructed by the simple Laurentia model shown in Fig. \ref{fig:Laurentia_paleolatitude} combined with the Phanerozoic model of \cite{Torsvik2017a} and as reconstructed by the model of \cite{Li2013a} for the Neoproterozoic. Proterozoic evaporite deposits in this panel are discussed in the text while Phanerozoic ones are taken from the compilation of \cite{Evans2006a}. The evaporite lines extend from the maximum to minimum age constraints with points at the preferred depositional age. The evaporite paleolatitude points are labeled with numbers that correspond to the numbers on the present-day location map. These numbers correspond to: 1 -- Altyn Formation (Belt Supergroup); 2: Wallace/Helena Formations -- Belt Supergroup; 3 -- Iqqittuq Formation (Bylot Supergroup); 4 -- Ten Stone Formation (Mackenzie Mountains Supergroup); 5 -- Minto Inlet Formation (Shaler Supergroup); 6 -- Redstone River Formation (Mackenzie Mountains Supergroup); 7 -- Kilian Formation (Shaler Supergroup); 8 -- Silurian Michigan Basin ; 9 -- Devonian Western Canada; 10 -- Carboniferous Canadian Maritime; 11 -- Carboniferous Sverdrup; 12 -- Permian Midcontinental USA; 13 -- Jurassic Gulf of Mexico.}
\label{fig:Laurentia_evaporites}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Grenville_Reconstruction.pdf}
\caption{\textbf{Paleogeographic reconstructions of Laurentia and other select Proterozoic continents leading up to Rodinia assembly in the late Mesoproterozoic modified from \cite{Swanson-Hysell2019a}}. The hypothesized connection between Siberia and Laurentia is implemented following \cite{Evans2016b} who interpret this relationship as persistent from 1.7 to 0.7 Ga. The reconstruction envisions the Kalahari and Amazonia cratons as conjugates with Laurentia in the Grenvillian orogeny. The Australia-East Antarctica relationship with Laurentia follows \cite{Swanson-Hysell2012a} and is similar to the Neoproterozoic reconstruction between the continents of \cite{Li2011a} and implementing their relative rotation of North Australia relative to South and West Australia. This configuration back to ca. 1140 Ma is consistent with a comparison between the Laurentia poles of that age and the coeval Mt. Isa dikes pole from North Australia and with the Keweenawan Track if the the Nonesuch and Freda poles are interpreted to be ca. 1080 (consistent with chronostratigraphic constraints; \citealp{Slotznick2018a}) with further motion by ca. 1070 Ma. The time slices show the rapid motion of Laurentia implied by the paleomagnetic poles which is consistent with the timing of collisional orogenesis associated with the Grenvillian orogeny.}
\label{fig:Grenville_reconstructions}
\end{figure*}

@@

The latest Mesoproterozoic to earliest Neoproterozoic pole path of the North China craton includes a swath of paleomagnetic poles with a similar arc length to the Keweenawan Track to Grenville Loop of Laurentia's APWP \citep{Zhao2019a, Zhang2021a}. While the chronostratigraphic age constraints on these North China poles are much looser than those from Laurentia, \cite{Zhao2019a} propose that the North China poles can be aligned with the Keweenawan Track to reconstruct North China as being conjoined to the northwest margin of Laurentia from prior to ca. 1110 Ma into the early Neoproterozoic. North China would have been at polar latitudes ca. 1110 Ma and moved rapidly with Laurentia as it transited towards the equator. A challenge with interpreting these North China poles as primary is that they include data from limestones that would reconstruct them to have been deposited at very high latitude ($>$80\textdegree\ for the limestones of the lower member of the Nanfen Formation). Such high-latitude limestones would require non-uniformitarian depositional conditions for carbonate precipitation which is more dominant in low latitudes due to the temperature dependence of carbonate saturation. As additional support for a North China--Laurentia connection, \cite{Zhao2019a} point to similarities in the detrital zircon age spectra between early Neoproterozoic sediments in NW Laurentia and North China basins as supporting this reconstruction. In particular, sediment transport from Laurentia could provide a source for ca. 1.18 Ga zircons (from the Shawinigan orogen) and ca. 1.08 Ga zircons (from the Grenville orogen). In the Laurentia basins, ca. 1.6 Ga zircons without a clear Laurentia source could be sourced from North China craton granites (e.g. \citealp{Wang2020a}). If North China was in this position, the timing of its arrival adjacent to Laurentia is unclear. The ca. 1220 Ma dikes pole of the North China craton is not coincident with the 1237 $\pm$ 5 Ma Sudbury dikes pole in this reconstructed position leading \cite{Zhao2019a} and \cite{Zhang2021a} to suggest that North China arrived on the Laurentian margin between ca. 1220 and 1110 Ma although they note a lack of evidence for known North China orogenesis at this time. Laurentia's pole path has a gap from the 1237 $\pm$ 5 Ma Sudbury pole to the 1184 $\pm$ 5 Ma KEEP EDITTING HERE pole when it needs to have asce In terms of departing from this position, one possibility is that its departure is associated with latest Neoproterozoic extension in northwest Laurentia.

%CHECK THE SUDBURY POINT FOR MYSELF

@@

Southern Siberia and northern Laurentia have been proposed to be connected on the basis of a similar history of Paleoproterozoic collision \citep{Rainbird1998a}, matches in the age of mafic intrusives rocks interpreted as shared large igneous provinces from the Paleoproterozoic to the Neoproterozoic \cite{Ernst2016a}, and comparisons of paleomagnetic poles \citep{Evans2011a, Evans2016b}. U-Pb dates from mafic intrusive rocks are interpreted by \cite{Ernst2016a} as resulting from shared large igneous provinces between southern Siberia and northern Laurentia from the time of Laurentia's amalgamation ca. 1.8 Ga all the way up to the time of the ca. 720 Ma Franklin large igneous province (LIP). For example, mafic intrusions and lavas that are grouped as the Irkutsk LIP in Siberia have been dated in Siberia to be similar in age to dates developed from the Franklin LIP \citep{Denyszyn2009a, Ernst2016a}. Comparisons of paleomagnetic poles between Laurentia and Siberia support this tight and internally stable fit between southern Siberia and northern Laurentia from ca. 1.64 to at least ca. 0.76 Ga  \citep{Evans2016b}. Overlap in Laurentia and Siberia pole positions with such a reconstruction are achieved in the Statherian Period of the Paleoproterozoic (the ca. 1.64 Ga Nersa complex of Siberia compared to the Melville Bugt dikes of Laurentia), the Calymmian Period of the Mesoproterozoic (the ca. 1.50 Ga Anabar intrusions with the ca. 1.48 Ga St. Francois Mountains igneous province of Laurentia), the Stentian Period of the Mesoproterozoic (a number of roughly chronostratigraphically constrained Siberia poles with the 1.11 to 1.08 Ga Keweenawan Track of Laurentia) and the Tonian Period of the Neoproterozoic (the ca.  Kitoi pole of Siberia with Tonian Laurentia poles).  The correlation of the Malgina pole of Siberia with the Keweenawan Track gains additional support in that it correlates the normal-polarity Maya superchron \citep{Gallet2012a} with the Portage Lake normal-polarity zone \citep{Swanson-Hysell2019a} that is interpreted as a normal-polarity superchron (termed the Keweenawan Normal Superchron in \citealp{Driscoll2016a}). This correlation both works with the tight fit and resolves hemispheric ambiguity to put both cratons together in the northern hemisphere ca. 1100 Ma (Fig. \ref{fig:Grenville_reconstructions}).

Connections between Archean provinces may have also existed prior to Laurentia assembly such as the hypothesized connection between the Slave Province and the Tungus Province of Siberia which has the Thelon orogen correlated to Paleoproterozoic orogenesis in the Akitkan fold belt \citep{Condie1994a, Rainbird1998a, Evans2011a}. The emplacement of the Franklin LIP ca. 720 Ma could have initiated break-up of a joint Laurentia-Siberia continent ending the long Proterozoic connection between the cratons.

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/baked_contact_timeline_all.pdf}
\caption{\textbf{Paleomagnetic poles with positive baked contact tests from Laurentia and other cratons}. This timeline shows the age of paleomagnetic poles with positive baked contact tests within the Nordic Paleogeography Workshop compilation \citep{Evans2021a}. Positive baked contact tests require the presence of an appreciable geomagnetic field. In turn, the presence of geomagnetic field requires heat flow across the core mantle boundary that is maintained by plate tectonics, but that would be stifled by a stagnant lid.}
\label{fig:baked_contact}
\end{figure*}

Even without considering other continents, there is strong evidence both in Laurentia's geological and paleomagnetic record for differential plate tectonic motion between 2.2 and 1.8 Ga. The continued history of accretionary orogenesis and the evaluation of Laurentia's pole path in comparison to other continents from 1.8 Ga onward supports the continual operation of plate tectonics throughout the rest of the Proterozoic and Phanerozoic as well. While this evidence fits with the majority of interpretations of the timing of initiation of modern-style plate tectonics (see summary in \citealp{Korenaga2013a}), there continue to be arguments proposing that a stagnant lid persisted through the Mesoproterozoic Era (1.6 to 1.0 Ga) and into the Neoproterozoic  with plate tectonics not initiating until ca. 0.8 Ga \citep{Hamilton2011a, Stern2018a}. These arguments rest largely on the relative lack of Proterozoic low-temperature high-pressure metamorphic rocks such as blueschists that form in subduction zones \citep{Stern2013a}. An alternative interpretation for this lack of blueschists in the Proterozoic is that such a shift in metamorphic regime is the predicted result of secular evolution of mantle chemistry rather than a harbinger of the onset of plate tectonics \citep{Palin2015a}. While this line of evidence is intriguing, to argue that there was not differential plate tectonic motion in the Paleoproterozoic and Mesoproterozoic is to ignore a vast breadth and depth of geological and paleomagnetic data. From a paleomagnetic perspective, there is strong support for independent and differential motion of the Slave and Superior provinces as is illustrated in Figure \ref{fig:Superior_Slave_recons}. From a geological perspective, the Trans-Hudson orogenic interval, the Grenville orogenic interval, and the Appalachian orogenic interval are all well-explained with a mobilistic interpretation that includes phases of accretionary orogenesis followed by collisional orogenesis (Fig. \ref{fig:tectonic_history}). One could counter that this perspective results from a plate-tectonic-centric viewpoint that lacks creativity to see the record as resulting from other processes than modern-style plate tectonics. However, in addition to the broad geological record showing an amalgamation of terranes as would be expected to arise through plate tectonics, there are also Proterozoic obducted ophiolites that are well-explained through accretionary plate tectonicsas as well as eclogites such as those preserved in the Trans-Hudson orogen \citep{Weller2017a}. These eclogites preserve evidence for high-pressure/low-temperature metamorphic conditions ca. 1.8 Ga. Similar to the Himalayan orogen, these rocks are interpreted to be the result of deep continental subduction and exhumation associated with convergent plate tectonics \citep{Weller2017a}. Outside of Laurentia, there are examples of Paleoproterozoic ecologites with geochemical affinity to oceanic crust such as that documented in the ca. 1.9 Ga Ubendian Belt of the Congo craton \citep{Boniface2012a}. Mesoproterozoic ophiolites were also obducted to Laurentia such as the Pyrites ophiolite complex within the Shawinigan orogen and the Coal Creek Domain of the Llano uplift \citep{Chiarenzelli2011a, McLelland2013a}.

>>>>

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Fig2_Tectonic_history.pdf}
\caption{\textbf{Simplified timeline of Laurentia's tectonic history over the past $\sim$1.8 billion years.} Brief summaries and references related to the orogenic and rifting episodes are given in the text. A timeline of large igneous provinces (LIPs) associated with typically brief and voluminous (or interpreted to be voluminous) volcanism is also shown. The interpreted age of paleomagnetic poles for Laurentia (not including separated terranes) compiled in this study for the Proterozoic and in \cite{Torsvik2012a} for the Phanerozoic is shown. Abbreviations on the figure: CAMP -- Central Atlantic Magmatic Province; proto-Cord -- proto-Cordilleran.}
\label{fig:tectonic_history}
\end{figure*}

@@

In northwest Laurentia from the Ogilvie Mountains of Yukon to Victoria Island, the sedimentary rock record is distinct from further south as it also records earlier Neoproterozoic basin development during the Tonian Period in addition to Cryogenian basin development \citep{Macdonald2012a}. Lithospheric extension is interpreted from basin development that accommodated deposition of the lower Fifteenmile Group with maximum depositional ages of ca. 1050 Ma with ongoing basin development ca. 812 Ma (age constraint from a U-Pb zircon date on a tuff within the upper Fifteenmile Group; \citealp{Macdonald2010a}) potentially through thermal subsidence \citep{Macdonald2012a}. Earlier basin development in the region recorded by the Mesoproterozoic/Neoproterozoic Pinguicula Group could provide valuable insight on tectonic history as it has been interpreted to have been deposited in an extensional basin \citep{Medig2016a}, however it is poorly constrained in terms of age --- older than the Fifteenmile Group and younger than the ca. 1382 Ma Hart River sills (which themselves have been interpreted to be emplaced in conjunction with rifting; \citealp{Verbaas2018a}). As in southwest Laurentia, extensional basin development continued into the Cryogenian and Ediacaran which accommodated deposition of the Windermere Supergroup \citep{Moynihan2019a}. Ediacaran rifting transitioned to thermal subsidence in the Cambrian and an early Paleozoic passive margin \citep{Moynihan2019a}. A similar history is interpreted from the North Slope terrane of Arctic Alaska (considered to have been part of northern Laurentia in the Neoproterozoic) where \cite{Strauss2019a} suggest that there was Tonian-Cryogenian extensional tectonism with limited overall lithospheric stretching that was followed by Ediacaran to early Cambrian extension that led to passive margin sedimentation.

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Fig3_Laurentia_poles_combined.pdf}
\caption{\textbf{Paleomagnetic poles from Laurentia.} Upper-left panel: Paleomagnetic poles from 1800 to 720 Ma for Laurentia (including Greenland). Portions of the apparent polar wander path (APWP) are referred to by the names Logan loop, Keweenawan track, and Grenville loop in the literature and those are labeled next to the associated poles. Upper-right panel:  Paleomagnetic poles from 1800 to 580 Ma for Laurentia (including those from the separated terranes of Greenland, Scotland and Svalbard rotated to Laurentia coordinates). The youngest poles from the Ediacaran Period have unusually variable positions as discussed in the text. Lower-left panel: Running mean APWP calculated with a 20 million year moving window. Lower-right panel: Poles for the Archean provinces of Laurentia prior to Laurentia's Paleoproterozoic amalgamation.}
\label{fig:Laurentia_poles}
\end{figure*}

@@

%The poles on this figure should be labeled
\begin{figure*}
\centering
\includegraphics[width=6 in]{../Figures/Fig4_Superior_Slave_reconstructions.pdf}
\caption{\textbf{Paleogeographic reconstructions developed using poles from the Superior, Slave and Rae provinces.} The polarity options that are chosen for the provinces are those that minimize total apparent polar wander path length. This model reconstructs a wide Manikewan ocean that underwent orthogonal closure rather than an alternative possibility of a narrower Manikewan ocean with a pivot-like closure. Paleomagnetic poles are shown colored to match their respective province with these provinces shown in present-day coordinates and labeled in the 0 Ma panel. Poles with ages that are within 25 million years of the given time slice are shown. The relatively well-resolved pole paths from the Superior and Slave provinces (Fig. \ref{fig:Laurentia_poles}) that are utilized for these reconstructions provide strong support for differential plate tectonic motion between 2220 and 1850 Ma.}
\label{fig:Superior_Slave_recons}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=0.9\textwidth]{../Figures/Fig5_Laurentia_paleolatitude.pdf}
\caption{\textbf{Laurentia paleolatitude through time in data and models.} Top panel: Paleolatitude for the city of Duluth on the southern margin of the Superior province (present-day coordinates of lat=46.79\textdegree N, lon=92.10\textdegree W) implied by paleomagnetic poles from Laurentia and associated terranes. The paleomagnetic poles are compiled in Table 2. Middle panel: Paleolatitude implied by Laurentia poles compared with that implied by published paleogeographic models and the simple Laurentia model used in this chapter for the reconstructions in Figure \ref{fig:Laurentia_reconstructions}. TC2017 refers to \cite{Torsvik2017a} and S-H 2017 refers to \cite{Swanson-Hysell2017a}. Bottom panel: the velocity implied by the continuous paleogeographic models in cm per year for the Duluth reference point on Laurentia.}
\label{fig:Laurentia_paleolatitude}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Fig6_Laurentia_reconstructions.pdf}
\caption{\textbf{Paleogeographic reconstructions of Laurentia at time intervals through the Proterozoic.} These reconstructions use the simple Laurentia pole interpolation model that is shown in Figure \ref{fig:Laurentia_paleolatitude} to reconstruct the tectonic elements of \cite{Whitmeyer2007a} shown in Figure \ref{fig:Laurentia_map}. Modern coastlines are maintained in these polygons so that the rotated orientations can be interpreted by the reader in comparison to Figure \ref{fig:Laurentia_map}. Paleomagnetic poles within 25 million years of each reconstruction time are plotted. All reconstructions have poles within such a time frame that provide constraints with the exception of the 850 Ma reconstruction which is shown faintly given this relative uncertainty in Laurentia's position.}
\label{fig:Laurentia_reconstructions}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=7 in]{../Figures/Fig7_Laurentia_evaporite_figure.pdf}
\caption{\textbf{Paleolatitude of Laurentia evaporites.} Left panel: The paleolatitude of evaporite deposits as reconstructed by the simple Laurentia model shown in Fig. \ref{fig:Laurentia_paleolatitude} combined with the Phanerozoic model of \cite{Torsvik2017a} and as reconstructed by the model of \cite{Li2013a} for the Neoproterozoic. Proterozoic evaporite deposits in this panel are discussed in the text while Phanerozoic ones are taken from the compilation of \cite{Evans2006a}. The evaporite lines extend from the maximum to minimum age constraints with points at the preferred depositional age. The evaporite paleolatitude points are labeled with numbers that correspond to the numbers on the present-day location map. These numbers correspond to: 1 -- Altyn Formation (Belt Supergroup); 2: Wallace/Helena Formations -- Belt Supergroup; 3 -- Iqqittuq Formation (Bylot Supergroup); 4 -- Ten Stone Formation (Mackenzie Mountains Supergroup); 5 -- Minto Inlet Formation (Shaler Supergroup); 6 -- Redstone River Formation (Mackenzie Mountains Supergroup); 7 -- Kilian Formation (Shaler Supergroup); 8 -- Silurian Michigan Basin ; 9 -- Devonian Western Canada; 10 -- Carboniferous Canadian Maritime; 11 -- Carboniferous Sverdrup; 12 -- Permian Midcontinental USA; 13 -- Jurassic Gulf of Mexico.}
\label{fig:Laurentia_evaporites}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Fig8_Grenville_Reconstruction.pdf}
\caption{\textbf{Paleogeographic reconstructions of Laurentia and other select Proterozoic continents leading up to Rodinia assembly in the late Mesoproterozoic modified from \cite{Swanson-Hysell2019a}}. The hypothesized connection between Siberia and Laurentia is implemented following \cite{Evans2016b} who interpret this relationship as persistent from 1.7 to 0.7 Ga. The reconstruction envisions the Kalahari and Amazonia cratons as conjugates with Laurentia in the Grenvillian orogeny. The Australia-East Antarctica relationship with Laurentia follows \cite{Swanson-Hysell2012a} and is similar to the Neoproterozoic reconstruction between the continents of \cite{Li2011a} and implementing their relative rotation of North Australia relative to South and West Australia. This configuration back to ca. 1140 Ma is consistent with a comparison between the Laurentia poles of that age and the coeval Mt. Isa dikes pole from North Australia and with the Keweenawan Track if the the Nonesuch and Freda poles are interpreted to be ca. 1080 (consistent with chronostratigraphic constraints; \citealp{Slotznick2018a}) with further motion by ca. 1070 Ma. The time slices show the rapid motion of Laurentia implied by the paleomagnetic poles which is consistent with the timing of collisional orogenesis associated with the Grenvillian orogeny.}
\label{fig:Grenville_reconstructions}
\end{figure*}

@@

The latest Mesoproterozoic to earliest Neoproterozoic pole path of the North China craton includes a swath of paleomagnetic poles with a similar arc length to the Keweenawan Track to Grenville Loop of Laurentia's APWP \citep{Zhao2019a, Zhang2021a}. While the chronostratigraphic age constraints on these North China poles are much looser than those from Laurentia, \cite{Zhao2019a} propose that the North China poles can be aligned with the Keweenawan Track to reconstruct North China as being conjoined to the northwest margin of Laurentia from prior to ca. 1110 Ma into the early Neoproterozoic. North China would have been at polar latitudes ca. 1110 Ma and moved rapidly with Laurentia as it transited towards the equator. A challenge with interpreting these North China poles as primary is that they include data from limestones that would reconstruct them to have been deposited at very high latitude ($>$80\textdegree\ for the limestones of the lower member of the Nanfen Formation). Such high-latitude limestones would require non-uniformitarian depositional conditions for carbonate precipitation which is more dominant in low latitudes due to the temperature dependence of carbonate saturation. As additional support for a North China--Laurentia connection, \cite{Zhao2019a} point to similarities in the detrital zircon age spectra between early Neoproterozoic sediments in NW Laurentia and North China basins as supporting this reconstruction. In particular, sediment transport from Laurentia could provide a source for ca. 1.18 Ga zircons (from the Shawinigan orogen) and ca. 1.08 Ga zircons (from the Grenville orogen). In the Laurentia basins, ca. 1.6 Ga zircons without a clear Laurentia source could be sourced from North China craton granites (e.g. \citealp{Wang2020a}). If North China was in this position, the timing of its arrival adjacent to Laurentia is unclear. The ca. 1220 Ma dikes pole of the North China craton is not coincident with the 1237 $\pm$ 5 Ma Sudbury dikes pole in this reconstructed position leading \cite{Zhao2019a} and \cite{Zhang2021a} to suggest that North China arrived on the Laurentian margin between ca. 1220 and 1110 Ma although they note a lack of evidence for known North China orogenesis at this time. Laurentia's pole path has a gap from the 1237 $\pm$ 5 Ma Sudbury pole to the 1184 $\pm$ 5 Ma during which it ascends to the Logan Loop (Figs. \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}). In its reconstructed position, the ca. 1220 Ma North China pole is close to this gap such that it is possible that the APWPs are consistent and that North China arrived on the northwest Laurentia margin prior to 1220 Ma. In terms of departing from this position, one possibility is that its departure is associated with latest Neoproterozoic extension in northwest Laurentia.

@@

Southern Siberia and northern Laurentia have been proposed to be connected on the basis of a similar history of Paleoproterozoic collision \citep{Rainbird1998a}, matches in the age of mafic intrusives rocks interpreted as shared large igneous provinces from the Paleoproterozoic to the Neoproterozoic \cite{Ernst2016a}, and comparisons of paleomagnetic poles \citep{Evans2011a, Evans2016b}. U-Pb dates from mafic intrusive rocks are interpreted by \cite{Ernst2016a} as resulting from shared large igneous provinces between southern Siberia and northern Laurentia from the time of Laurentia's amalgamation ca. 1.8 Ga all the way up to the time of the ca. 720 Ma Franklin large igneous province (LIP). For example, mafic intrusions and lavas that are grouped as the Irkutsk LIP in Siberia have been dated in Siberia to be similar in age to dates developed from the Franklin LIP \citep{Denyszyn2009a, Ernst2016a}. Comparisons of paleomagnetic poles between Laurentia and Siberia support this tight and internally stable fit between southern Siberia and northern Laurentia from ca. 1.64 to at least ca. 0.76 Ga  \citep{Evans2016b}. Overlap in Laurentia and Siberia pole positions with such a reconstruction are achieved in the Statherian Period of the Paleoproterozoic (the ca. 1.64 Ga Nersa complex of Siberia compared to the Melville Bugt dikes of Laurentia), the Calymmian Period of the Mesoproterozoic (the ca. 1.50 Ga Anabar intrusions with the ca. 1.48 Ga St. Francois Mountains igneous province of Laurentia), the Stentian Period of the Mesoproterozoic (a number of roughly chronostratigraphically constrained Siberia poles with the 1.11 to 1.08 Ga Keweenawan Track of Laurentia) and the Tonian Period of the Neoproterozoic (the ca.  Kitoi pole of Siberia with Tonian Laurentia poles).  The correlation of the Malgina pole of Siberia with the Keweenawan Track gains additional support in that it correlates the normal-polarity Maya superchron \citep{Gallet2012a} with the Portage Lake normal-polarity zone \citep{Swanson-Hysell2019a} that is interpreted as a normal-polarity superchron (termed the Keweenawan Normal Superchron in \citealp{Driscoll2016a}). This correlation both works with the tight fit and resolves hemispheric ambiguity to put both cratons together in the northern hemisphere ca. 1100 Ma (Fig. \ref{fig:Grenville_reconstructions}). Connections between Archean provinces may have also existed prior to Laurentia assembly such as the hypothesized connection between the Slave Province and the Tungus Province of Siberia which has the Thelon orogen correlated to Paleoproterozoic orogenesis in the Akitkan fold belt \citep{Condie1994a, Rainbird1998a, Evans2011a}.

The interpretation that the Franklin LIP is recorded in both Laurentia and Siberia constrains the break-up between the continents to post-date 720 Ma. While there could have been some extension in the Tonian Period and at the time of the Franklin emplacement, the record of northern Laurentia subsidence indicates Ediacaran to early Cambrian extensional tectonism (e.g. \citealp{Strauss2019a}). Therefore, it might not have been until the late Ediacaran Period that the long-term Proterozoic connection between the cratons ended.

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Fig9_baked_contact_timeline_all.pdf}
\caption{\textbf{Paleomagnetic poles with positive baked contact tests from Laurentia and other cratons}. This timeline shows the age of paleomagnetic poles with positive baked contact tests within the Nordic Paleogeography Workshop compilation \citep{Evans2021a}. Positive baked contact tests require the presence of an appreciable geomagnetic field. In turn, the presence of geomagnetic field requires heat flow across the core mantle boundary that is maintained by plate tectonics, but that would be stifled by a stagnant lid.}
\label{fig:baked_contact}
\end{figure*}

Even without considering other continents, there is strong evidence both in Laurentia's geological and paleomagnetic record for differential plate tectonic motion between 2.2 and 1.8 Ga. The continued history of accretionary orogenesis and the evaluation of Laurentia's pole path in comparison to other continents from 1.8 Ga onward supports the continual operation of plate tectonics throughout the rest of the Proterozoic and Phanerozoic as well. While this evidence fits with the majority of interpretations of the timing of initiation of modern-style plate tectonics (see summary in \citealp{Korenaga2013a}), there continue to be arguments proposing that a stagnant lid persisted through the Mesoproterozoic Era (1.6 to 1.0 Ga) and into the Neoproterozoic  with plate tectonics not initiating until ca. 0.8 Ga \citep{Hamilton2011a, Stern2018a}. These arguments rest largely on the relative lack of Proterozoic low-temperature high-pressure metamorphic rocks such as blueschists that form in subduction zones \citep{Stern2013a}. An alternative interpretation for this lack of blueschists in the Proterozoic is that such a shift in metamorphic regime is the predicted result of secular evolution of mantle chemistry rather than a harbinger of the onset of plate tectonics \citep{Palin2015a}. While this line of evidence is intriguing, to argue that there was not differential plate tectonic motion in the Paleoproterozoic and Mesoproterozoic is to ignore a vast breadth and depth of geological and paleomagnetic data. From a paleomagnetic perspective, there is strong support for independent and differential motion of the Slave and Superior provinces as is illustrated in Figure \ref{fig:Superior_Slave_recons}. From a geological perspective, the Trans-Hudson orogenic interval, the Grenville orogenic interval, and the Appalachian orogenic interval are all well-explained with a mobilistic interpretation that includes phases of accretionary orogenesis followed by collisional orogenesis (Fig. \ref{fig:tectonic_history}). One could counter that this perspective results from a plate-tectonic-centric viewpoint that lacks creativity to see the record as resulting from other processes than modern-style plate tectonics. However, in addition to the broad geological record showing an amalgamation of terranes as would be expected to arise through plate tectonics, there are also Proterozoic obducted ophiolites that are well-explained through accretionary plate tectonics as as well as eclogites such as those preserved in the Trans-Hudson orogen \citep{Weller2017a}. These eclogites preserve evidence for high-pressure/low-temperature metamorphic conditions ca. 1.8 Ga. Similar to the Himalayan orogen, these rocks are interpreted to be the result of deep continental subduction and exhumation associated with convergent plate tectonics \citep{Weller2017a}. Outside of Laurentia, there are examples of Paleoproterozoic ecologites with geochemical affinity to oceanic crust such as that documented in the ca. 1.9 Ga Ubendian Belt of the Congo craton \citep{Boniface2012a}. Mesoproterozoic ophiolites were also obducted to Laurentia such as the Pyrites ophiolite complex within the Shawinigan orogen and the Coal Creek Domain of the Llano uplift \citep{Chiarenzelli2011a, McLelland2013a}.
